---
title: "Optimization with GP"
subtitle: |
date: '`r Sys.Date()`'
author: Nam-Anh Tran
format:
  html: 
    theme: darkly
    toc: false
fontsize: 12pt
reference-location: margin
citation-location: margin
number-sections: true
fig-cap-location: bottom
fig-height: 6
fig-width: 8.5
link-citations: true
linkcolor: deeppink
indent: true
classoption: [onecolumn, portrait]
bibliography: 03_references.bib
csl: ../vancouver.csl
editor: source
editor_options: 
  chunk_output_type: console
execute: 
  eval: true
  warning: false
---

\renewcommand{\bf}[1]{\boldsymbol{#1}}

```{r}
if(!library(pacman, logical.return = T)) install.packages("pacman")
pacman::p_load(tidyverse, plgp, laGP, lhs, mvtnorm, plotly, knitr, kableExtra)
```

## Motivations

We aim to optimize the function $f(x)$, i.e.

$$
x_* = \arg\max_{x \in \mathcal X}f(x)
$$

The traditional solution proceeds by evaluating the function $f$ and taking derivative. Many variants of this framework has been proposed to overcome the challenges of taking derivative, since this step sometimes is not a piece of cake. This impasse can sometimes appears at the step of function evaluation, i.e. the structure of $f$ is a blackbox and, thus, taking derivative is impossible. Thus, the Bayesian optimization appears as a solution for this problem.  

## Basic examples

Let consider the following function which does not have a closed-form. Note that examples in this section can be found in Gramacy et al.[@gramacy2020surrogates] 

```{r}
#| label: fig-01
#| fig-cap: "The surface of the objective function."


obj_func <- function(X)
{
if(is.null(nrow(X))) X <- matrix(X, nrow=1)
m <- 8.6928
s <- 2.4269
x1 <- 4*X[,1] - 2
x2 <- 4*X[,2] - 2

a <- 1 + (x1 + x2 + 1)^2 *
(19 - 14*x1 + 3*x1^2 - 14*x2 + 6*x1*x2 + 3*x2^2)
b <- 30 + (2*x1 - 3*x2)^2 *
(18 - 32*x1 + 12*x1^2 + 48*x2 - 36*x1*x2 + 27*x2^2)
f <- log(a*b)
f <- (f - m)/s
return(f)
}
#--------------------

xx = seq(0,1, 0.01)
xx2d = expand.grid(xx,xx)
yy = obj_func(xx2d)
yymat = matrix(yy, nrow = length(xx), byrow = TRUE)

fig1<-
  plot_ly(x = xx,y = xx, z = yymat, type = "surface", colors = c("gray50"),
        showscale = F, opacity = .7)

fig1
```

```{r}
#| label: fig-02
#| fig-cap: "The contour of the objective function."

fig2<-
  plot_ly(x = xx,y = xx, z = yymat, type = "contour",
        showscale = F, opacity = .7, colors = "gray", contours = list(showlabels = TRUE))
fig2
```


```{r}
#| warning: false
#| label: fig-03
#| fig-cap: "Contour and the initial design."

set.seed(4444)
ninit <- 20
X <- randomLHS(ninit, 2)
y <- obj_func(X)
####
fig2|> add_trace(x = X[,1], y = X[,2],z = yymat, type = "scatter", mode = "markers", marker = list(color = "blue"), showlegend = F)
```

```{r}
#| label: fig-04
#| fig-cap: The red dots are the optimal desing points.

da <- darg(list(mle=TRUE, max=0.5), randomLHS(1000, 2))
gpi <- newGPsep(X, y, d=da$start, g=1e-6, dK=TRUE)
mleGPsep(gpi, param="d", tmin=da$min, tmax=da$max, ab=da$ab)$msg
####
obj.mean <- function(x, gpi) predGPsep(gpi, matrix(x, nrow=1), lite=TRUE)$mean
####

opt_ls<- c()
  
repeat {
m <- which.min(y)
opt <- optim(X[m,], obj.mean, lower=0, upper=1,
method="L-BFGS-B", gpi=gpi)
ynew <- obj_func(opt$par)
opt_ls<- rbind(opt_ls, c(opt$par, ynew))
if(abs(ynew - y[length(y)]) < 1e-4) break;
updateGPsep(gpi, matrix(opt$par, nrow=1), ynew)
mle <- mleGPsep(gpi, param="d", tmin=da$min, tmax=da$max, ab=da$ab)
X <- rbind(X, opt$par)
y <- c(y, ynew)
}
deleteGPsep(gpi)

plot_ly(x = xx,y = xx, z = yymat, type = "contour",
        showscale = F, opacity = .7, contours = list(showlabels = TRUE), colors = "gray")|> 
  add_trace(x = X[,1], y = X[,2], type = "scatter", mode = "markers", marker = list(color = "blue"), showlegend = F)|>
  add_trace(x = opt_ls[,1], y = opt_ls[,2], type = "scatter", mode = "lines", marker = list(color = "red"), showlegend = F)
```


```{r}
#| column: margin
#| label: tbl-01
#| tbl-cap: this is

compose(as_tibble, data.frame)(opt_ls)|>
  rename(Y = "X3")|>
  round(5)|>
  arrange(Y)|>
  knitr::kable()|>
  kable_styling(bootstrap_options = c("striped", "hover"))|>
  row_spec(0, background = "deeppink")|>
   scroll_box(height = "300px")
  
```






